### 2023-07-12 tex/conc.tex
[ writing ] various up to WRel example

In this thesis, I have developed a foundation for semiring-annotated calculi in
natural deduction style.

FIXME: this chapter is directly from the ESOP22 paper.

@@

\begin{itemize}
  \item Novel simultaneous substitution for linear type theories
  \item Generalise \citet{AACMM21} work to substructural logics
  \item Is this a reusable library? Yes, examples
  \item Fitch-style systems (can't do).
  \item Methodology for metatheory of new calculi
  \item Semirings have been a reasonably robust basis.
  \item Can the code be directly generalised?
    Abstract away the algebraic structure of contexts.
\end{itemize}

\section{Addendum: (lack of) partiality}\label{sec:part}
\input{part.tex}

>>>>

In this thesis, I have developed a foundation for semiring-annotated calculi
presented in natural deduction style.
I have given a consolidated account of the semiring-annotated calculus $\name$,
including its relations to existing linear and modal calculi.
As part of this, I have adapted what I call \emph{bunched connectives} from
\cite{RPKV20} as a way to state the typing rules of the calculus as well as to
work with the metatheory.
The distinction between sharing and separating conjunction given by the bunched
connectives corresponds well to the notions of \emph{additive} and
\emph{multiplicative} connectives in the object language, respectively.
Following this, I have given a novel linear algebra-based definition of
\emph{environment} for semiring-annotated calculi, together with a motivation
which may serve as a basis for the corresponding definition for other
substructural systems.
The adequacy of this definition of environment is shown by my implementation of
simultaneous renaming and substitution, as well as other operations on
environments, like composition of renamings and substitutions.

With the details of $\name$ worked out, I then moved on to adapting the work of
\citet{AACMM21} so as to make it able to capture semiring-based usage
restrictions, as found in $\name$.
The syntax descriptions of the resulting system are based on the bunched
connectives, and are shown to be expressive enough to encode calculi of a
variety of forms.
In adapting the semantics, I am forced to be precise about \emph{sharing} and
\emph{separating} bunched connectives, but largely add these as a refinement of
the work of \citet{AACMM21}.
I provide the renaming and substitution operations for all expressible calculi.
I also provide more specialised examples of semantic traversals:
a usage elaborator, an NbE algorithm, a denotational semantics, and translations
between different calculi.
The usage elaborator gives an unexpected example of generic programming, which
one could not write without syntax descriptions.
% Other examples...
Together, these examples show the applicability and versatility of the framework
I have developed.

FIXME: the next two paragraphs are directly from the ESOP22 paper.

@@

\begin{itemize}
  \item Novel simultaneous substitution for linear type theories
  \item Generalise \citet{AACMM21} work to substructural logics
  \item Is this a reusable library? Yes, examples
  \item Methodology for metatheory of new calculi
  \item Semirings have been a reasonably robust basis.
  \item Can the code be directly generalised?
    Abstract away the algebraic structure of contexts.
\end{itemize}

\section{Future work}

\begin{itemize}
  \item Fitch-style systems (can't do).
\end{itemize}

\subsection{Further work}

\paragraph{Equality}
Perhaps the most fundamental missing piece from the metatheoretic account of
semiring-annotated calculi I have given in this thesis is equations between
terms.
Reasoning about equality between terms and environments is a problem I have
tried to solve, but I have not arrived at a satisfactory solution in the time
available to me.

I believe that the basic difficulty of giving an account for equality in a
linear setting is the proliferation of $\Sigma$-types.
For example, describing equality between two applications of
\TirName{$\with$-I} is immediate: $\Gamma \vdash (M, N) = (M', N') : A \with B$
if and only if $\Gamma \vdash M = M' : A$ and $\Gamma \vdash N = N' : B$.
However, to do the same with \TirName{$\otimes$-I} requires us to be careful
about the contexts of the subterms.
The two applications of \TirName{$\otimes$-I} may a priori split the context in
different ways, and should only be equated when those splittings are equal
(in the appropriate sense).
If the splittings are equal, then the contexts of the subterms will line up, and
only then can the subterms themselves be compared for equality.
These multiple stages come about because $\plr{T * U}\,\Gamma$ is a
$\Sigma$-type, and equality of elements of $\Sigma$-types always follows this
pattern.
Such reasoning becomes even more complex for environments, which are equivalent
to iterated $*$-families.
Additionally, it is unclear what effect subsumption of contexts (like subusaging
in this thesis or explicit structural rules in other calculi) should have on
equality, particularly when the subsumption commutes with parts of the subterms.

Previous work on linear logic has preferred to deal with equations only on
untyped terms. \todo{Check this.}

\paragraph{Polymorphism}
An important feature of most contemporary statically typed programming languages
is polymorphism.
In particular, parametric polymorphism over types can be used to significantly
improve code reuse, and is well undestood theoretically via System $F$ and its
variants.
I have not considered polymorphism in this thesis, and neither did
\citet{AACMM21} in their paper, so whether it can be supported in the framework
presented earlier is an open question.
However, \citet{Autosubst15} did apply their related system to System $F_{<:}$,
so it may be possible to support polymorphic systems using their techniques.

A separate but related question concerns polymorphism over usage annotations.
The status of polymorphism over usage annotations is less well established both
in practice and in theory.
\Citet{Granule18} present an implementation allowing for polymorphism of usage
annotations, and even polymorphism over semirings, but provide no more than
example programs to justify the feature.
This thesis provides no advance on understanding annotation polymorphism, unless
it can be encoded into a semiring to fit the framework.

%\subsection{Further generalisations}

\paragraph{Structure of contexts}
As I have presented it, the work of \citet{AACMM21} has two axes in which it is
generic: the syntax, which can be controlled through syntax descriptions to
produce a wide range of calculi and features; and the semantics, where we can
produce a wide range of maps out of terms with the help of environments.
To this, the work of this thesis has added a third axis of genericity: the usage
discipline of variables, as described by a partially ordered semiring.

Starting at least with the bunched connectives in \cref{sec:lnd}, if not earlier
when talking about usage contexts forming modules over the semiring of
annotations, I have made productive use of abstractions over the basic usage
annotations throughout this thesis.
These abstractions suggest a next step of completely abstracting away much of
the representation of contexts and their individual entries.
One may imagine that it is possible to develop a framework in which the required
operations and properties of contexts are axiomatised, similar to how usage
annotations are axiomatised to form a semiring in this thesis, and to how
categories-with-families models are defined~\citep{Dybjer95,CCD19}.
Instances of such a framework would include the work of \citet{RPKV20}, which
uses a very similar bunched connective abstraction over a very different
representation of contexts, based on relational interleaving of lists.

The use of semirings is motivated in this thesis and elsewhere largely because
they are general enough to cover a wide range of examples.
However, I cannot claim to have a derivation from first principles of why we
should choose partially ordered semirings over any of a range of similar
algebraic structures.
Additionally, some of the specific constructions done in this thesis fit
somewhat unnaturally with the semiring-based approach.
For example, when translating semiring-annotated systems to traditional systems,
I tended to need to make a \emph{bottom-up} assumption
(\cref{def:DILL-bottom-up,def:PD-bottom-up}) so as to avoid some ``junk'' facts
given by the semiring.
Meanwhile, the usage elaborator of \cref{sec:usage-elaborator} eschews the
``forward'' computation of semiring operations in favour of non-deterministic
backwards computation, e.g., from a sum to the collection of possible summands.
Possibly consciously working more abstractly, as described in the previous
paragraph, would make a more natural structure appear.

If we are to retain an annotation-based approach to usage restrictions, then a
possible feature request that falls out of the encoding of linear/non-linear
logic is to have some sort of kinding system by which different kinds of types
are annotated using different sets of annotations.
In the L/nL example, we would want linear types to be annotated with $\gr0$ and
$\gr1$, and intuitionistic types to be annotated with $\gr\omega$ (as the sole
element of a trivial instance of an algebraic structure), with no crossover
between the two kinds.
Algebraic means to handle such mixed-kind usage vectors may be inspired by the
work of \citet{Hart95,MF21} on dimensional analysis in linear algebra.

\subsubsection{Partiality}\label{sec:part}
\input{part.tex}

### 2023-07-12 tex/examples.tex
[ writing ] various up to WRel example

@@

Standard denotational semantics falls out as a somewhat trivial case of
\AgdaRecord{Semantics}.
A lot of work is done in the generic traversal \AgdaFunction{semantics} to
maintain a $\V$-environment, where $\V$ is a reasonably variable-like semantic
family.

{\color{red}Old stuff below\ldots}

To justify the name \emph{semantics}, we give an example traversal that is a
denotational semantics in the usual sense.
The semantics we take is a refinement of that of \citet{AbelBernardy2020},
which gives a way to extract parametricity theorems from substructurally typed
programs.
Example theorems are that all linear terms act as permutations on some fixed
set of resources, and all monotonically typed terms are really monotonic in the
way the typing suggests they are.

To abbreviate this section, we use a simplified syntax compared to \name{}.
We allow for an arbitrary family of base types \AgdaBound{BaseTy}, and a single
type former \mbox{\ExecuteMetaData[\WReltex]{rAToB}}, equivalent to
\mbox{\ExecuteMetaData[\PaperExamplestex]{BangrAToB}} from the earlier system.

\ExecuteMetaData[\WReltex]{Ty}

In the term syntax, $\lambda$-abstraction now binds a variable with annotation
\AgdaBound{r}, while application needs to scale its argument by \AgdaBound{r}
(both in accordance with the function type they are acting on).

\ExecuteMetaData[\WReltex]{AnnArr}

As a running example, we take the usage annotations to be the 4-element
variance posemiring (\cref{thm:variance}).
We establish the property that all terms are monotonic in their free variables.
This monotonicity can be covariant or contravariant (or neither or both)
depending on the annotation of each free variable.
This provides an additional example to those of \citeauthor{AbelBernardy2020}.

We will take semantics of this system into
\emph{world-indexed relations}~\cite{AbelBernardy2020,context-constrained}.
A world-indexed relation (\AgdaRecord{WRel}) over a poset of worlds
\AgdaBound{W} is a set over which
we have a \AgdaBound{W}-indexed binary relation satisfying a presheaf-like
property with respect to the order on \AgdaBound{W}.

@@

Morphisms (\AgdaRecord{WRelMor}) between world-indexed relations \AgdaBound{R}
and \AgdaBound{S} consist of a mapping between the underlying sets such that, at
each fixed world \AgdaBound{w}, the mapping preserves relatedness from
\AgdaBound{R} to \AgdaBound{S}.

@@

When the poset of worlds forms a (relational) commutative monoid, such
world-indexed relations support a symmetric monoidal closed structure, with
objects denoted \AgdaFunction{I$^R$},
\AgdaFunction{\AgdaUnderscore{}$\otimes^R$\AgdaUnderscore{}}, and
\AgdaFunction{\AgdaUnderscore{}$\multimap^R$\AgdaUnderscore{}},.
These reuse the bunched connectives \AgdaRecord{$I^*$}, \AgdaRecord{$\sep$}, and
\AgdaRecord{$\wand$}, now over worlds rather than contexts.

The final piece of semantics we need is a \emph{bang} operator.
We allow the
semantic \emph{bang} to be an arbitrary annotation-indexed functor on
world-indexed relations.
This functor must respect all of the structure on the indices, making it a
graded comonad over multiplication, as well as being lax monoidal at any
particular index \AgdaBound{r}.

@@

\begin{example}
  With \AgdaBound{W} being the 1-element set and annotations coming from the
  variance semiring, we can define the following \emph{bang}.
  It is always the identity on the set component, while the relation component
  consists of flipping the relation for contravariance and taking conjunctions
  to achieve both covariance and contravariance.
  When we want neither covariance nor contravariance, we use the always true
  predicate on worlds \AgdaFunction{$\dot1$}.

@@

The semantics of a type is given by \AgdaFunction{$\llbracket$\_$\rrbracket$},
which maps into world-indexed relations.
The function type is interpreted using \AgdaFunction{$\multimap^R$} and
\AgdaField{!$^R$}.
Contexts are interpreted by \AgdaFunction{$\llbracket$\_$\rrbracket^c$}, using
\AgdaFunction{$\otimes^R$} and \AgdaFunction{I$^R$}.
Terms are interpreted as morphisms by the open family
\AgdaFunction{$\llbracket$\_$\vdash$\_$\rrbracket$}.
Variables are interpreted by \AgdaFunction{lookup$^R$} (definition omitted).

\ExecuteMetaData[\WReltex]{lookupR-type}

Now we give a \AgdaRecord{Semantics}.
The choice of \AgdaBound{$\V$} as
\AgdaRecord{\AgdaUnderscore{}$\sqni$\AgdaUnderscore{}} is somewhat arbitrary,
given that a standard denotational semantics would not use intermediate
environments in the same sense as renaming and substitution, but it allows us to
reuse the standard facts that variables support renaming and identity
environments.
With this choice of \AgdaBound{$\V$} and \AgdaBound{$\C$}, we interpret
environment entries by \AgdaFunction{lookup$^R$}.
Meanwhile, for the logical rules, we ignore environments by using
\AgdaFunction{reify} to just deal with morphisms in an extended context.
As such, $\lambda$-abstractions are easy to interpret, while applications
require some massaging to remove the extension by an empty context, followed by
some plumbing to split the interpretation of the context according to the usage
constraints and feed the interpretation of the argument \AgdaBound{n$'$} into
the interpretation of the function \AgdaBound{m$'$}.

@@

>>>>

\section{Normalisation by evaluation}\label{sec:nbe}

To give an example of a common operation on programming languages, I present a
normalisation by evaluation (NbE) algorithm~\citep{BS91} for a fragment of
$\name$.
The algorithm contains not only a structural induction on a term using an
environment, but also functions which create terms.
It therefore gives an example of various parts of the framework.

Though it is easy to define a syntax of normal and neutral forms, I do not do so
here.
Converting between terms and normal/neutral forms introduces significant
overheads, and the syntax of normal/neutral forms is somewhat complicated by the
invariant that all variables are of neutral type.
It would be worthwhile future work to work out this refinement, thereby showing
that the normaliser presented in this section actually puts terms into normal
form.
Another limitation is that I do no reasoning about equivalence of terms, so
there is no proof that the resultant term is equivalent to the original.

For the sake of simplicity and conciseness, I consider just the fragment
containing a unary function type, but whose argument carries a usage annotation.
The partially ordered semiring of usage annotations is arbitrary throughout.
I also include an arbitrary set of base types \AgdaBound{BaseTy} via constructor
\AgdaInductiveConstructor{$\iota$}.

\ExecuteMetaData[\NbESimpletex]{Ty}

With only function types, the only term formers are application and
$\lambda$-abstraction.
These are similar to the corresponding term formers in $\name$, but application
scales its argument in accordance with the usage annotation of the function
type, and $\lambda$-abstraction binds a variable with the given usage
annotation.
For convenience, I define an alias \AgdaFunction{Term} for the open family of
terms in this type system of unspecified (\AgdaPostulate{$\infty$}) size.

\ExecuteMetaData[\NbESimpletex]{system}

The NbE model of types is given by \AgdaFunction{\_$\vDash$\_}.
It is defined by induction on the type.
The base type is interpreted by terms of base type, while
the function type is interpreted by the Kripke function space (essentially a
special case of \AgdaFunction{Kripke} with $\Delta = \gr rA$, though reusing
the \AgdaFunction{Kripke} definition causes difficulty for the termination
checker).
The interpretation of the base type additionally requires a \AgdaRecord{Lift}
of its universe level, because \AgdaFunction{$\Box^r$} produces a large type
(a type in \AgdaPrimitive{Set$_1$}).
The largeness of \AgdaFunction{$\Box^r$} comes from the largeness of
environments, and hence renamings, because of the relationally specified linear
map.

\ExecuteMetaData[\NbESimpletex]{vDash}

\AgdaFunction{\_$\vDash$\_} is a renameable family, as can be seen by cases on
the type.
It is renameable at the base type because terms are renameable, and it is
renameable at function type because any open type formed using
\AgdaFunction{$\Box^r$} is renameable by composition of renamings.
I refer to this proof of renameability as
\AgdaFunction{ren\textasciicircum$\vDash$}.

Now that I am working with a specific syntax, I introduce the function
\AgdaFunction{reify[]}.
This is a variant of the \AgdaFunction{reify} function given in \cref{sec:reify}
usable only for rules that bind no new variables --- i.e., the parameter
\AgdaBound{$\Delta$} to \AgdaFunction{Kripke} is the empty context.

\begin{lemma}[\AgdaFunction{reify[]}]
  We have a function of type
  $\mathrm{Kripke}\,\V\,\C\,\plr{\cdot} \rightarrowtriangle \C$ for any open
  families $\V$ and $\C$.
\end{lemma}
\begin{proof}
  Let $b : \mathrm{Kripke}\,\V\,\C\,\plr{\cdot}\,\Gamma\,A$.
  We give $b$ the identity renaming (corresponding to no extension of the
  context), yielding
  $b\,1 : \plr{\plr{-} \env\V {\cdot} \wand \plr{-} \sdtstile{}\C A}\,\Gamma$.
  By \cref{thm:construct-env}, we have that $\plr{-} \env\V {\cdot}$ is
  equivalent to $I^*$, and by monoidal closure,
  $\plr{I^* \wand \plr{-} \sdtstile{}\C A}$ is equivalent to
  $\plr{-} \sdtstile{}\C A$, so the result follows.
\end{proof}

The first part of the NbE algorithm I give is the evaluator.
The evaluator interprets terms into the model via an environment of values from
the model, making it a clear instance of the generic semantic traversal.
Specifically, I fix both \AgdaBound{$\V$} and \AgdaBound{$\C$} to
\AgdaFunction{\_$\vDash$\_} (allowing the conversion from semantic values to
semantic terms to be given by the identity function \AgdaFunction{id}), and
provide the other required data as below.
The case for $\lambda$-abstraction is almost trivial, given that its job is to
convert the Kripke function \AgdaBound{M} coming from the generic traversal into
a value in the model of function type (i.e.\ a Kripke function).
The main thing we need to do is to wrap the argument to the function into a
singleton environment using \AgdaFunction{[-]$^e$}.
In the case for function application, we use \AgdaFunction{reify[]} to get rid of
the extraneous Kripke functions coming from the generic traversal.
Having done so for both subterms, we apply the value corresponding to
\AgdaBound{M} (which is a Kripke function) to the value corresponding to
\AgdaBound{N} in an unextended context (\AgdaFunction{1$^r$}), with usage
information divided up as in the original application.

\ExecuteMetaData[\NbESimpletex]{evalSem}

The function \AgdaFunction{eval} is given by the appropriate instantiation of
the generic traversal \AgdaFunction{semantics}.
Note that we have not yet shown that \AgdaFunction{\_$\vDash$\_} supports
identity environments (which requires the reify and reflect functions), so we
cannot yet evaluate arbitrary open terms.

\ExecuteMetaData[\NbESimpletex]{eval}

The final pieces of NbE are the \emph{reify} and \emph{reflect} functions.
I give these the names \AgdaFunction{nbeReify} and \AgdaFunction{nbeReflect}
respectively to disambiguate from the \AgdaFunction{reify} and
\AgdaFunction{reify[]} functions of the framework.
\AgdaFunction{nbeReify} and \AgdaFunction{nbeReflect} are defined mutually by
recursion on the type, as is standard in NbE.
Therefore, neither uses the generic semantic traversal, though both make use of
some of the general components I have developed in this thesis.

At base type, values in the model are terms (modulo \AgdaRecord{Lift}), so the
conversion between these and terms is trivial.
At function type, \AgdaFunction{nbeReify} is given a Kripke function
\AgdaBound{v} and aims to turn it into a normal form of function type.
It does so in the canonical way --- introducing a $\lambda$-abstraction --- and
then aims to produce a term in an extended context.
Because of the context extension, we apply \AgdaBound{v} using the renaming
\AgdaFunction{$\swarrow^r$}\AgdaSpace{}\AgdaSymbol{:}\AgdaSpace{}%
\AgdaBound{$\Gamma$}\AgdaSpace{}\AgdaFunction{++$^c$}\AgdaSpace{}%
\AgdaFunction{[}\AgdaSpace{}\AgdaField{0\#}\AgdaSpace{}\AgdaFunction{$\bullet$}%
\AgdaSpace{}\AgdaBound{A}\AgdaSpace{}\AgdaFunction{]$^c$}
to get it ``back'' to the original context.
The splitting \AgdaBound{sp+} adds together $\Gamma, \gr0A$ and
$\gr0\Gamma, \gr rA$ to get $\Gamma, \gr rA$, while \AgdaBound{sp*} scales down
$\gr0\Gamma, \gr rA$ to $\gr0\Gamma, \gr1A$.
Semantic value \AgdaBound{v} then wants a semantic value of type $A$, which it
gets from reflecting the newly bound variable into the model.
Finally, \AgdaBound{v} produces a semantic value, which a recursive call to
\AgdaFunction{nbeReify} at the smaller type \AgdaBound{B} converts into a term
(in fact, a normal form).
Meanwhile, \AgdaFunction{nbeReflect} at function type is given a syntactic
function (in fact, a neutral form) and aims to turn it into a Kripke function.
To produce a Kripke function, we consider a renaming \AgdaBound{$\rho$} into
\AgdaBound{$\Gamma$}, splittings of the resultant context, and a semantic value
\AgdaBound{N} of type \AgdaBound{A}.
To produce the value of type \AgdaBound{B}, we recursively call
\AgdaFunction{nbeReflect} at smaller type \AgdaBound{B}, but with a larger term
formed by reifying \AgdaBound{N} (again of smaller type) and applying it to the
term \AgdaBound{M}.
To make the application work, we must rename both \AgdaBound{M} and
\AgdaBound{N} to ignore the (empty) context extensions introduced by the
application rule, and we must also rename \AgdaBound{M} by \AgdaBound{$\rho$} to
put \AgdaBound{M} into the extended context the reflection is being done at.

\ExecuteMetaData[\NbESimpletex]{reify-reflect}

Finally, we compose \AgdaFunction{eval}, \AgdaFunction{nbeReify}, and
\AgdaFunction{nbeReflect} to get the normaliser \AgdaFunction{nbe}.
The definition of \AgdaFunction{nbe} uses \AgdaFunction{eval} to interpret the
term into the model, and then uses \AgdaFunction{nbeReify} to read the semantic
value back as a term (a normal form).
However, as noted earlier, we need to give an environment of type
$\Gamma \env\vDash \Gamma$ to \AgdaFunction{eval} in order to interpret an open
term in its own context.
I provide this environment via \cref{thm:env-id}, which says that if we have a
mapping from variables in $\Gamma$ to values in $\Gamma$, we can get the desired
identity environment.
The required map is given in
\AgdaFunction{identityEnv\textasciicircum{}$\vDash$}
by turning the variable into a term using \AgdaInductiveConstructor{`var} and
then applying \AgdaFunction{nbeReflect}.
The definition \AgdaFunction{identityEnv\textasciicircum{}$\vDash$} is marked as
an \AgdaKeyword{instance} so that it is picked up by
\AgdaFunction{id\textasciicircum{}Env}.

\ExecuteMetaData[\NbESimpletex]{nbe}

The main limitation of this example is that the results are not specified as
being in normal form by their type.
Ideally, we would have a syntax of normal and neutral forms for
\AgdaFunction{nbe} and supporting functions to target.
We can produce such a syntax --- similarly to how \citet{AACMM21} produce a
syntax for bidirectional type checking --- but this syntax is difficult to work
with because of the crucial restriction on the allowable kind of variables ---
namely that variables can only appear as neutral forms, and not normal forms.
I believe that this difficulty could be worked around via carefully placed
assertions that various contexts are well formed --- and in particular that the
extended context introduced by a \AgdaFunction{$\Box^r$} is well formed.

Also, I have not given an account of many of the connectives of $\name$.
Of particular interest would be the types with pattern-matching eliminators ---
$I$, $\otimes$, $0$, $\oplus$, and $\oc$.
I believe these could be handled by ordinary techniques for NbE for sums ---
probably not doing $\eta$-expansion for these types.

@@

Standard denotational semantics falls out as a somewhat trivial case of
\AgdaRecord{Semantics}.
A lot of work is done in the generic traversal \AgdaFunction{semantics} to
maintain a $\V$-environment, where $\V$ is a reasonably variable-like semantic
family.
For a simple denotational semantics, we set $\V$ to be variables

As an example denotational semantics for a semiring-annotated calculus, I take
a semantics in world-indexed relations.
This is a refinement of the semantics given by \citet{AbelBernardy2020},
which gives a way to extract parametricity theorems from substructurally typed
programs.
Example theorems are that all linear terms act as permutations on some fixed
set of resources, and all monotonically typed terms are really monotonic in the
way the typing suggests they are.

For the sake of brevity, I use the same term language as in \cref{sec:nbe}, with
a set of base types and function types with a usage-annotated domain type.

As a running example, I take the usage annotations to be the 4-element
variance posemiring (\cref{def:monotonicity-posemiring}).
I establish the property that all terms are monotonic in their free variables.
This monotonicity can be covariant or contravariant (or neither or both)
depending on the annotation of each free variable (respectively,
\AgdaInductiveConstructor{$\uparrow\uparrow$},
\AgdaInductiveConstructor{$\downarrow\downarrow$},
\AgdaInductiveConstructor{$\wn\wn$}, and
\AgdaInductiveConstructor{\textasciitilde\textasciitilde}).
This provides an additional example to those of \citet{AbelBernardy2020}.

In the semantics of this section, types are interpreted as
\emph{world-indexed relations}~\citep{AbelBernardy2020,context-constrained}.
A world-indexed relation (\AgdaRecord{WRel}) over a poset of worlds
\AgdaBound{W} is a set \AgdaField{set} over which
we have a \AgdaBound{W}-indexed binary relation \AgdaField{ren} satisfying a
presheaf-like property \AgdaBound{subres} with respect to the order on
\AgdaBound{W}.
I declare the record type \AgdaRecord{WRel} to have
\AgdaKeyword{no-eta-equality} so that the usual $\eta$-law of records does not
apply to it.
The lack of an $\eta$-law allows the connectives on \AgdaRecord{WRel}
(e.g.\ \AgdaFunction{\_$\otimes^R$\_} and \AgdaFunction{\_$\multimap^R$\_},
defined below) to be definitionally injective, helping the unification in Agda's
elaborator.

@@

Morphisms (\AgdaRecord{WRelMor}) between world-indexed relations \AgdaBound{R}
and \AgdaBound{S} consist of a mapping \AgdaField{set$\Rightarrow$} between the
underlying sets such that, at each world \AgdaBound{w}, the mapping preserves
relatedness from \AgdaBound{R} to \AgdaBound{S}.
In the type of the field \AgdaField{rel$\Rightarrow$}, the world index is
handled implicitly via \AgdaFunction{\_$\rightarrowtriangle$\_}.

@@

When the poset of worlds forms a (relational) commutative monoid, such
world-indexed relations support a symmetric monoidal closed structure, with
objects denoted \AgdaFunction{I$^R$},
\AgdaFunction{\AgdaUnderscore{}$\otimes^R$\AgdaUnderscore{}}, and
\AgdaFunction{\AgdaUnderscore{}$\multimap^R$\AgdaUnderscore{}},.
These reuse the bunched connectives \AgdaRecord{$I^*$}, \AgdaRecord{$\sep$}, and
\AgdaRecord{$\wand$}, now over worlds rather than contexts.
Their definitions are listed below.

\ExecuteMetaData[\WReltex]{IR}
\ExecuteMetaData[\WReltex]{tensorR}
\ExecuteMetaData[\WReltex]{lollyR}

The final piece of semantics we need is a \emph{bang} operator.
I allow the
semantic \emph{bang} to be an arbitrary annotation-indexed functor on
world-indexed relations.
This functor must respect all of the structure on the indices, making it a
graded comonad over multiplication, as well as being lax monoidal at any
particular index \AgdaBound{r}.

@@

The properties required of the semantic \emph{bang} operator are slightly
weaker than those given by \citet{AbelBernardy2020}.
Specifically, the morphisms given by \AgdaField{!$^R$-1}, \AgdaField{!$^R$-I},
and \AgdaField{!$^R$-$\otimes$} only go one way, rather than being
bi-implications as they are for \citet{AbelBernardy2020}.
\citeauthor{AbelBernardy2020} need the stronger laws because of their strong
eliminator for tensor products, discussed in \cref{sec:alt}.
Also, I do not use the \emph{non-idempotent intersection} operators (which
handle worlds in the same way as \AgdaFunction{I$^R$} and
\AgdaFunction{\AgdaUnderscore{}$\otimes^R$\AgdaUnderscore{}}, but keep the
underlying set the same, rather than using products of the underlying sets),
and thus do not have axioms for them.
I avoid the need for non-idempotent intersections by giving the semantics all at
once, rather than first giving a $\mathrm{Set}$-semantics and then giving a
$\mathrm{WRel}$-semantics on top of it.

\begin{example}
  With \AgdaBound{W} being the 1-element set and annotations coming from the
  variance semiring, we can define the following \emph{bang}.
  It is always the identity on the set component, while the relation component
  consists of flipping the relation for contravariance and taking conjunctions
  to achieve both covariance and contravariance.
  When we want neither covariance nor contravariance, we use the always true
  predicate on worlds \AgdaFunction{$\dot1$}.
  With the worlds being trivial, so too is the property \AgdaField{subres}.

@@

Let us assume that we have a mapping
\AgdaBound{$\upiota\llbracket$\_$\rrbracket$} from base types to world-indexed
relations.
We extend this interpretation to all types via
\AgdaFunction{$\llbracket$\_$\rrbracket$}, shown below, with
the function type being interpreted using \AgdaFunction{$\multimap^R$} and
\AgdaField{!$^R$}.
Contexts are interpreted by \AgdaFunction{$\llbracket$\_$\rrbracket^c$}, using
\AgdaFunction{$\otimes^R$} and \AgdaFunction{I$^R$}, and \AgdaField{!$^R$}
for singleton contexts.
Terms are interpreted as morphisms by the open family
\AgdaFunction{$\llbracket$\_$\vdash$\_$\rrbracket$}.
Variables are interpreted by \AgdaFunction{lookup$^R$} (definition omitted).

\ExecuteMetaData[\WReltex]{sem}
\ExecuteMetaData[\WReltex]{sem-vdash}
\ExecuteMetaData[\WReltex]{lookupR-type}

The laws \AgdaField{!$^R$-0}, \AgdaField{!$^R$-+}, and \AgdaField{!$^R$-*} lift
from singleton contexts to all contexts via lemmas \AgdaFunction{ctx-0},
\AgdaFunction{ctx-+}, and \AgdaFunction{ctx-*}, with the latter saying that we
have morphisms of type
\mbox{$\sem{\plr{\gr r\grP}\gamma} \to \oc^R\gr r\sem{\grP\gamma}$}.
We can see each of these three lemmas as ways to turn the algebraic structure of
contexts into structured world-indexed relations --- allowing us to use general
facts about \AgdaFunction{I$^R$}, \AgdaFunction{\_$\otimes^R$\_}, and
\AgdaField{!$^R$}, respectively.

Now I give a \AgdaRecord{Semantics}.
The choice of \AgdaBound{$\V$} as
\AgdaRecord{\AgdaUnderscore{}$\sqni$\AgdaUnderscore{}} is somewhat arbitrary,
given that a standard denotational semantics would not use intermediate
environments in the same sense as renaming and substitution, but it allows us to
reuse the standard facts that variables support renaming and identity
environments.
With this choice of \AgdaBound{$\V$} and \AgdaBound{$\C$}, we interpret
environment entries by \AgdaFunction{lookup$^R$}.

Meanwhile, for the logical rules, we ignore environments by using
\AgdaFunction{reify} and \AgdaFunction{reify[]} to just deal with morphisms in
an extended context.
The crucial structure of world-indexed relations is given by lemmas
\AgdaFunction{curry$^R$} and \AgdaFunction{uncurry$^R$}, which translate between
\mbox{\AgdaRecord{WRelMor}\AgdaSpace{}\AgdaSymbol(\AgdaBound{R}\AgdaSpace{}%
\AgdaFunction{$\otimes^R$}\AgdaSpace{}\AgdaBound{S}\AgdaSymbol)%
\AgdaSpace{}\AgdaBound{T}} and
\mbox{\AgdaRecord{WRelMor}\AgdaSpace{}\AgdaBound{R}\AgdaSpace{}\AgdaSymbol(%
\AgdaBound{S}\AgdaSpace{}\AgdaFunction{$\multimap^R$}\AgdaSpace{}\AgdaBound{T}%
\AgdaSymbol)}.
Also, \AgdaFunction{map-$\otimes^R$} implements the tensor
product of world-indexed relation morphisms.
With these lemmas, the interpretations of $\lambda$-abstraction and application
are straightforward.
For $\lambda$-abstraction, we have
\mbox{$\sem{\Gamma, \gr rA} = \sem\Gamma \otimes^R \oc^R\gr r\sem{A}$},
so we just need to use \AgdaFunction{curry$^R$} to get the desired semantic
function in the original context.
For application, we are producing the composition pictured in
\cref{fig:Wrel-app}.
In this picture, I ignore the possible subusaging between $\grP + \gr r\grQ$ and
the original context for the sake of notational suggestiveness.

@@

\begin{figure}
  \centering
  \begin{tikzpicture}
    \node(0) {$\sem{\plr{\grP + \gr r\grQ}\gamma}$};
    \node(1) [right=of 0] {$\otimes$};
    \node(2) [above=1em of 1] {$\sem{\grP\gamma}$};
    \node(3) [below=1em of 1] {$\sem{\gr r\grQ\gamma}$};
    \node(4) [right=of 3] {$\oc\gr r\sem{\grQ\gamma}$};
    \node(5) [right=of 4] {$\oc\gr r\sem{A}$};
    \node(6) [above=1em of 5] {$\otimes$};
    \node(7) [above=1em of 6] {$\sem{\grP\gamma}$};
    \node(8) [right=of 6] {$\sem{B}$};

    \draw[->] (0) to node[above] {\AgdaFunction{ctx-+}} (1);
    \draw[->] (2) to node[above] {\AgdaFunction{id$^R$}} (7);
    \draw[->] (3) to node[below] {\AgdaFunction{ctx-*}} (4);
    \draw[->] (4) to node[below] {\AgdaBound{n}} (5);
    \draw[->] (6) to node[above] {\AgdaBound{m}} (8);
  \end{tikzpicture}
  \caption{Interpretation of application in the world-indexed relation
    semantics}
  \label{fig:Wrel-app}
\end{figure}

### 2023-07-12 tex/gen-sem.tex
[ writing ] various up to WRel example

If we want to capture a broader range of traversals, including not just
syntactic but also \emph{semantic} operations, we must be able to target things
other than terms, and act in an interesting way on term constructors.
Doing a straight generalisation of the type of \AgdaFunction{trav}, this
suggests that we want a function with the following type, where
\AgdaBound{$\C$} is the type family we are targeting.

@@

However, we want to treat binding constructs specially, particularly because
there are semanticses with no notion of binding.
We instead provide a function from values to computations that works in any
\emph{extension} of the current context.
Keeping \AgdaField{$\swarrow^k$} as before, we get the following semantic
replacement for \AgdaRecord{Kit}.

>>>>

If we want to capture a broader range of traversals, including not just
syntactic but also \emph{semantic} operations, we must be able to target things
other than terms, and act in an interesting way on term constructors.
\Citet{ACMM17} show how to do this generalisation of \AgdaFunction{trav} to a
function \AgdaFunction{sem} with the following type, where
\AgdaBound{$\C$} is the type family we are targeting.
In this section, I discuss the assumptions needed to implement such a function.

@@

However, we want to treat binding constructs specially, particularly because
there are semantics with no notion of binding.
We instead provide a function from values to computations that works in any
\emph{extension} of the current context.
Keeping \AgdaField{$\swarrow^k$} as before, we get the following semantic
replacement for \AgdaRecord{Kit}.
